\bf The forestecology R package for fitting and assessing nieghborhood models of the effect of interspecific competition on the growth of trees

### 2021-03-08 paper/forestecology.Rmd
My last edits

---
title: The forestecology R package for fitting and assessing models of interspecies competitive effects on the growth of trees

@@

abstract: 
- |
  | 1. Neighborhood competition models are powerful tools to measure the effect of interspecific competition. Statistical methods to ease the application of these models are currently lacking.
  | 2. We present the `forestecology` package providing methods to i) specify neighborhood competition models, ii) evalulate the effect of competitor species identity using permutation tests, and iii) measure model performance using spatial cross-validation.  Following @allen_permutation_2020, we implement a Bayesian linear regression neighborhood competition model.
  | 3. We demonstrate the package's functionality using data from the Smithsonian Conservation Biology Institute's large forest dynamics plot, part of the ForestGEO global network of reseach sites. Given ForestGEO's data collection protocols and data formatting standards, the package was designed with cross-site compatibility in mind. We demonstrate that both i) competitor species identity matters and ii) that not spatially cross-validating leads to error estimates that are overly optimistic. 
  | 4. The package features i) `tidyverse`-like structure whereby verb-named functions can be modularly "piped" in sequence, ii) functions with standardized inputs/outputs of simple features `sf` package class, and iii) an S3 object-oriented implementation of the Bayesian linear regression model. These three facts allow for clear articulation of all the steps in the sequence of analysis and easy wrangling and visualization of the geospatial forestry data. Furthermore, while the package only has Bayesian linear regression implemented, the package was designed with extensibility to other methods in mind.

keywords:
- forest ecology, interspecific competition, neigbhorhood competition, tree growth, R, ForestGEO, cross-validation, spatial statistics,

@@

Repeat-censused forest plots offer excellent data to test neighborhood models of the effect of competition on the growth of trees (@canham_neighborhood_2004). Neighborhood models of competition have been used to: test whether the species identity of a competitor matters (@uriarte_spatially_2004); measure species-specific competition coefficients (@das_effect_2012 @tatsumi_estimating_2016); test competing models to see what structures competitive interactions, e.g. traits or phylogeny (@allen_permutation_2020 @uriarte_trait_2010); and inform selective logging practices (@canham_neighborhood_2006). Although these are well-described methods, no methods are currently available for their easy application. Here we address this in an R package. We largely follow the methods presented in @allen_permutation_2020. The package is written to model stem radial growth between two censuses based on neighborhood competition. It assumes a single large plot using the ForestGEO plot data structure as a guide (@andersonteixeira_ctfs-forestgeo_2015).

@@

The resulting RMSE of `r model_rmse_shuffle` based on the permutation test is `r ifelse(model_rmse_shuffle < model_rmse, "smaller", "larger")` than the earlier RMSE of `r model_rmse`, suggesting that models that do incorporate competitor species identity better fit the data. We conduct an analysis of the full SCBI plot in Section \ref{results} below.

@@

The resulting RMSE of `r model_rmse_cv` computed using cross-validation is `r ifelse(model_rmse_cv > model_rmse, "larger", "smaller")` than the earlier RMSE of `r model_rmse`, suggesting that models that do not take the inherent spatial autocorrelation of the data into account generate error estimates that are overly optimistic; in our case RMSE's that are too low. We conduct an analysis of the full SCBI plot in Section \ref{results} below.

# Results {#results}

Figure \ref{fig:scbi-simulation} highlights the importance of spatial cross validating when interpreting results from neighborhood compeition models. Without cross-validation the competitor idenity did matter, the non-permuted competitor species had a much lower RMSE than the permuted one. But once we include the spatial cross validation this improvement disaprees. So in this 9 ha subplot of the SCBI plot competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other places (@allen_permutation_2020 @uriarte_spatially_2004).

```{r scbi-simulation, out.width="100%", fig.cap="Root mean squared error of models for standard, permuted, and spatial cross-validated error estimates.", echo = FALSE}
include_graphics("simulation_results/2021-03-03_scbi_49_shuffles.png")
```

@@

The `forestecology` package provides an accessible way to fit and test models of neighborhood competition. Currently it is written to work with data from ForestGEO plots. But it could easily be modified to work on any single large, mapped forest plot in which at least two measurements of each individual have been taken. With a little bit of work the package could also be applied to forest inventory data in which several small plots are spatially separated, e.g., USFS Forest Inventory. In a future version of the package we also hope to make it possible to model plant mortality in addition to plant growth. The package follows the guidelines for `tidy` data, `sf` spatial structure., and S3 open-oriented model structure. These make for ease of use. We hope that the package will further the use of neighborhood competition models to better understand what structures plant competition.

>>>>

---
title: The forestecology R package for fitting and assessing nieghborhood models of the effect of interspecific competition on the growth of trees

@@

abstract: 
- |
  | 1. Neighborhood competition models are powerful tools to measure the effect of interspecific competition. Statistical methods to ease the application of these models are currently lacking.
  | 2. We present the `forestecology` package providing methods to i) specify neighborhood competition models, ii) evalulate the effect of competitor species identity using permutation tests, and iii) measure model performance using spatial cross-validation.  Following @allen_permutation_2020, we implement a Bayesian linear regression neighborhood competition model.
  | 3. We demonstrate the package's functionality using data from the Smithsonian Conservation Biology Institute's large forest dynamics plot, part of the ForestGEO global network of reseach sites. Given ForestGEO's data collection protocols and data formatting standards, the package was designed with cross-site compatibility in mind. We highlight the importance of spatial cross validation when interpreting model results. 
  | 4. The package features i) `tidyverse`-like structure whereby verb-named functions can be modularly "piped" in sequence, ii) functions with standardized inputs/outputs of simple features `sf` package class, and iii) an S3 object-oriented implementation of the Bayesian linear regression model. These three facts allow for clear articulation of all the steps in the sequence of analysis and easy wrangling and visualization of the geospatial forestry data. Furthermore, while the package only has Bayesian linear regression implemented, the package was designed with extensibility to other methods in mind.

keywords:
- forest ecology, interspecific competition, neigbhorhood competition, tree growth, R, ForestGEO, spatial cross validation

@@

Repeat-censused forest plots offer excellent opportunities to test neighborhood models of the effect of competition on the growth of trees (@canham_neighborhood_2004). Neighborhood models of competition have been used to: test whether the species identity of a competitor matters (@uriarte_spatially_2004); measure species-specific competition coefficients (@das_effect_2012 @tatsumi_estimating_2016); test competing models to see what structures competitive interactions, e.g. traits or phylogeny (@allen_permutation_2020; @uriarte_trait_2010); and inform selective logging practices (@canham_neighborhood_2006). Although these are well-described methods, no methods are currently available for their easy application. Here we address this in an R package. We largely follow the methods presented in @allen_permutation_2020. The package is written to model stem radial growth between two censuses based on neighborhood competition.

@@

The resulting RMSE of `r model_rmse_shuffle` based on the permutation test is `r ifelse(model_rmse_shuffle < model_rmse, "smaller", "larger")` than the earlier RMSE of `r model_rmse`, suggesting that models that do incorporate competitor species identity better fit the data.

@@

The resulting RMSE of `r model_rmse_cv` computed using cross-validation is `r ifelse(model_rmse_cv > model_rmse, "larger", "smaller")` than the earlier RMSE of `r model_rmse`, suggesting that models that do not take the inherent spatial autocorrelation of the data into account generate error estimates that are overly optimistic; in our case RMSE's that are too low.

# Importance of spatial cross validation

The `run_cv` function also accepts the `run_shuffle` argument. This permutes the competitor species, as described above, but does so when calculating predicted growth with the cross validated scheme. Figure \ref{fig:scbi-simulation} compares model performance when permuting competitor species and calculating RMSE with and without cross validation. Without cross-validation the competitor identity did matter, the non-permuted competitor species had a much lower RMSE than the permuted one. But once we include the spatial cross validation this improvement disappears. So in this 9 ha subplot of the SCBI plot competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other places (@allen_permutation_2020 @uriarte_spatially_2004). This highlights the importance of cross validation, without it the model was overfit.

```{r scbi-simulation, out.width="100%", fig.cap="Root mean squared error of models for standard, permuted, and spatial cross-validated error estimates. The dotted lines show non-permuted competitor identity, while the histogras so the RMSE for 49 permutations. The colors indicate whether cross validaton was used.", echo = FALSE}
include_graphics("simulation_results/2021-03-03_scbi_49_shuffles.png")
```

@@

The `forestecology` package provides an accessible way to fit and test models of neighborhood competition. Currently it is written to work with data from ForestGEO plots. But it could easily be modified to work on any single large, mapped forest plot in which at least two measurements of each individual have been taken. With a little bit of work the package could also be applied to forest inventory data in which several small plots are spatially separated, e.g., USFS Forest Inventory. In a future version of the package we also hope to make it possible to model plant mortality in addition to plant growth. The package follows the guidelines for `tidy` data, `sf` spatial structure, and S3 open-oriented model structure. We hope that the package will increase the use of neighborhood competition models to better understand what structures plant competition.

### 2021-03-08 paper/forestecology.tex
My last edits

\if0\blind
{
  \title{\bf The forestecology R package for fitting and assessing models of
interspecies competitive effects on the growth of trees}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf The forestecology R package for fitting and assessing models of
interspecies competitive effects on the growth of trees}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
1. Neighborhood competition models are powerful tools to measure the
effect of interspecific competition. Statistical methods to ease the
application of these models are currently lacking.\\
2. We present the \texttt{forestecology} package providing methods to i)
specify neighborhood competition models, ii) evalulate the effect of
competitor species identity using permutation tests, and iii) measure
model performance using spatial cross-validation. Following
\citet{allen_permutation_2020}, we implement a Bayesian linear
regression neighborhood competition model.\\
3. We demonstrate the package's functionality using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO global network of reseach sites. Given ForestGEO's
data collection protocols and data formatting standards, the package was
designed with cross-site compatibility in mind. We demonstrate that both
i) competitor species identity matters and ii) that not spatially
cross-validating leads to error estimates that are overly optimistic.\\
4. The package features i) \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence, ii)
functions with standardized inputs/outputs of simple features
\texttt{sf} package class, and iii) an S3 object-oriented implementation
of the Bayesian linear regression model. These three facts allow for
clear articulation of all the steps in the sequence of analysis and easy
wrangling and visualization of the geospatial forestry data.
Furthermore, while the package only has Bayesian linear regression
implemented, the package was designed with extensibility to other
methods in mind.
\end{abstract}

\noindent%
{\it Keywords:} forest ecology, interspecific competition, neigbhorhood competition,
tree growth, R, ForestGEO, cross-validation, spatial statistics,
\vfill

@@

Repeat-censused forest plots offer excellent data to test neighborhood
models of the effect of competition on the growth of trees
(\citet{canham_neighborhood_2004}). Neighborhood models of competition
have been used to: test whether the species identity of a competitor
matters (\citet{uriarte_spatially_2004}); measure species-specific
competition coefficients (\citet{das_effect_2012}
\citet{tatsumi_estimating_2016}); test competing models to see what
structures competitive interactions, e.g.~traits or phylogeny
(\citet{allen_permutation_2020} \citet{uriarte_trait_2010}); and inform
selective logging practices (\citet{canham_neighborhood_2006}). Although
these are well-described methods, no methods are currently available for
their easy application. Here we address this in an R package. We largely
follow the methods presented in \citet{allen_permutation_2020}. The
package is written to model stem radial growth between two censuses
based on neighborhood competition. It assumes a single large plot using
the ForestGEO plot data structure as a guide
(\citet{andersonteixeira_ctfs-forestgeo_2015}).

@@

The resulting RMSE of 0.131 based on the permutation test is larger than
the earlier RMSE of 0.128, suggesting that models that do incorporate
competitor species identity better fit the data. We conduct an analysis
of the full SCBI plot in Section \ref{results} below.

@@

The resulting RMSE of 0.14 computed using cross-validation is larger
than the earlier RMSE of 0.128, suggesting that models that do not take
the inherent spatial autocorrelation of the data into account generate
error estimates that are overly optimistic; in our case RMSE's that are
too low. We conduct an analysis of the full SCBI plot in Section
\ref{results} below.

\hypertarget{results}{%
\section{Results}\label{results}}

Figure \ref{fig:scbi-simulation} highlights the importance of spatial
cross validating when interpreting results from neighborhood compeition
models. Without cross-validation the competitor idenity did matter, the
non-permuted competitor species had a much lower RMSE than the permuted
one. But once we include the spatial cross validation this improvement
disaprees. So in this 9 ha subplot of the SCBI plot competitive
interactions do not depend on the identity of the competitor, which is
the opposite of what has been observed in other places
(\citet{allen_permutation_2020} \citet{uriarte_spatially_2004}).

@@

\caption{Root mean squared error of models for standard, permuted, and spatial cross-validated error estimates.}\label{fig:scbi-simulation}
\end{figure}

@@

The \texttt{forestecology} package provides an accessible way to fit and
test models of neighborhood competition. Currently it is written to work
with data from ForestGEO plots. But it could easily be modified to work
on any single large, mapped forest plot in which at least two
measurements of each individual have been taken. With a little bit of
work the package could also be applied to forest inventory data in which
several small plots are spatially separated, e.g., USFS Forest
Inventory. In a future version of the package we also hope to make it
possible to model plant mortality in addition to plant growth. The
package follows the guidelines for \texttt{tidy} data, \texttt{sf}
spatial structure., and S3 open-oriented model structure. These make for
ease of use. We hope that the package will further the use of
neighborhood competition models to better understand what structures
plant competition.

>>>>

\if0\blind
{
  \title{\bf The forestecology R package for fitting and assessing nieghborhood
models of the effect of interspecific competition on the growth of trees}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf The forestecology R package for fitting and assessing nieghborhood
models of the effect of interspecific competition on the growth of trees}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
1. Neighborhood competition models are powerful tools to measure the
effect of interspecific competition. Statistical methods to ease the
application of these models are currently lacking.\\
2. We present the \texttt{forestecology} package providing methods to i)
specify neighborhood competition models, ii) evalulate the effect of
competitor species identity using permutation tests, and iii) measure
model performance using spatial cross-validation. Following
\citet{allen_permutation_2020}, we implement a Bayesian linear
regression neighborhood competition model.\\
3. We demonstrate the package's functionality using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO global network of reseach sites. Given ForestGEO's
data collection protocols and data formatting standards, the package was
designed with cross-site compatibility in mind. We highlight the
importance of spatial cross validation when interpreting model
results.\\
4. The package features i) \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence, ii)
functions with standardized inputs/outputs of simple features
\texttt{sf} package class, and iii) an S3 object-oriented implementation
of the Bayesian linear regression model. These three facts allow for
clear articulation of all the steps in the sequence of analysis and easy
wrangling and visualization of the geospatial forestry data.
Furthermore, while the package only has Bayesian linear regression
implemented, the package was designed with extensibility to other
methods in mind.
\end{abstract}

\noindent%
{\it Keywords:} forest ecology, interspecific competition, neigbhorhood competition,
tree growth, R, ForestGEO, spatial cross validation
\vfill

@@

Repeat-censused forest plots offer excellent opportunities to test
neighborhood models of the effect of competition on the growth of trees
(\citet{canham_neighborhood_2004}). Neighborhood models of competition
have been used to: test whether the species identity of a competitor
matters (\citet{uriarte_spatially_2004}); measure species-specific
competition coefficients (\citet{das_effect_2012}
\citet{tatsumi_estimating_2016}); test competing models to see what
structures competitive interactions, e.g.~traits or phylogeny
(\citet{allen_permutation_2020}; \citet{uriarte_trait_2010}); and inform
selective logging practices (\citet{canham_neighborhood_2006}). Although
these are well-described methods, no methods are currently available for
their easy application. Here we address this in an R package. We largely
follow the methods presented in \citet{allen_permutation_2020}. The
package is written to model stem radial growth between two censuses
based on neighborhood competition.

@@

The resulting RMSE of 0.131 based on the permutation test is larger than
the earlier RMSE of 0.128, suggesting that models that do incorporate
competitor species identity better fit the data.

@@

The resulting RMSE of 0.14 computed using cross-validation is larger
than the earlier RMSE of 0.128, suggesting that models that do not take
the inherent spatial autocorrelation of the data into account generate
error estimates that are overly optimistic; in our case RMSE's that are
too low.

\hypertarget{importance-of-spatial-cross-validation}{%
\section{Importance of spatial cross
validation}\label{importance-of-spatial-cross-validation}}

The \texttt{run\_cv} function also accepts the \texttt{run\_shuffle}
argument. This permutes the competitor species, as described above, but
does so when calculating predicted growth with the cross validated
scheme. Figure \ref{fig:scbi-simulation} compares model performance when
permuting competitor species and calculating RMSE with and without cross
validation. Without cross-validation the competitor identity did matter,
the non-permuted competitor species had a much lower RMSE than the
permuted one. But once we include the spatial cross validation this
improvement disappears. So in this 9 ha subplot of the SCBI plot
competitive interactions do not depend on the identity of the
competitor, which is the opposite of what has been observed in other
places (\citet{allen_permutation_2020} \citet{uriarte_spatially_2004}).
This highlights the importance of cross validation, without it the model
was overfit.

@@

\caption{Root mean squared error of models for standard, permuted, and spatial cross-validated error estimates. The dotted lines show non-permuted competitor identity, while the histogras so the RMSE for 49 permutations. The colors indicate whether cross validaton was used.}\label{fig:scbi-simulation}
\end{figure}

@@

The \texttt{forestecology} package provides an accessible way to fit and
test models of neighborhood competition. Currently it is written to work
with data from ForestGEO plots. But it could easily be modified to work
on any single large, mapped forest plot in which at least two
measurements of each individual have been taken. With a little bit of
work the package could also be applied to forest inventory data in which
several small plots are spatially separated, e.g., USFS Forest
Inventory. In a future version of the package we also hope to make it
possible to model plant mortality in addition to plant growth. The
package follows the guidelines for \texttt{tidy} data, \texttt{sf}
spatial structure, and S3 open-oriented model structure. We hope that
the package will increase the use of neighborhood competition models to
better understand what structures plant competition.
